Modeling the effect of aneuploidy on cancer evolution

### 2023-06-04 EvolutionaryRescue1.tex
going over todos with RS and started revising mean time

% line numbers
\usepackage[displaymath, mathlines]{lineno}
\renewcommand\linenumberfont{\normalfont\small\sffamily}
%\linenumbers
\modulolinenumbers[2]

@@

% TODO remove the commented out paragraphs if you dont need them. OK

@@

%%%%%%%%%%%
\begin{figure}[!t]
\centering
\includegraphics[width=0.75\textwidth]{Figures/figureAneuploidy.pdf}
\caption{
\textbf{Model illustration.}
% TODO in the figure change labels to Wildtype, Aneuploid, Mutant (instead of Euploid etc)
\textbf{(A)} A population of cancer cells is composed of wildtype, aneuploid, and mutant cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively. 
Wildtype cells can become aneuploid at rate $u$. Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$. Color denotes the relative growth rates of the three genotypes such that $\lambda_w - \mu_w < \lambda_a - \mu_a < \lambda_m - \mu_m$. \textbf{(B)} The wildtype and the mutant are susceptible and resistant, respectively, to the drug. The aneuploid may be tolerant, non-growing, or or partially resistant.
}
\label{figureAneuploidy}
\end{figure}
%%%%%%%%%%%

@@

The state of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the state (\Cref{figureAneuploidy}):
\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\quad\left(\text{birth of wildtype cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
(-1,+1,0)&:\quad uw_t\quad\left(\text{wildtype cell becomes aneuploid}\right),\\
(-1,0,+1)&:\quad vw_t\quad\left(\text{wildtype cell becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,-1,+1)&:\quad va_t\quad\left(\text{aneuploid cell becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_am_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_am_t\quad\left(\text{death of mutant cell}\right).
\end{flalign*}
\end{subequations}
Each iteration of the simulation loop starts with computing the rates $\nu_j$ of each event $j$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $j$ is $p_j=\nu_j/\sum_i \nu_i$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$. % TODO more details - you need to specify the events and how they change the number of cells. OK
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
\begin{equation*}
m_t>\left\lfloor-\frac{3\log10}{\log\left(\frac{\mu_m}{\lambda_m}\right)}\right\rfloor+1,
\end{equation*}

\paragraph{$\tau$-leaping.}
% TODO "is large" - how large?
When the size of the initial population is large we utilize $\tau$-leaping \citep{gillespie2001approximate}, where change in number of cells of genotype $i$ in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_i\Delta t$.
If the increment is negative and larger then the subpopulation size then updated to be zero. % TODO "and larger..."??

@@

In our analysis we assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources.
A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
To simulate birth and death rates that depend on the number of cells in the population, we transform the rates of division and death to the following:
\begin{align*}
\lambda_w' &= \lambda_w, \\
\mu_w' &= \mu_w,\\
\lambda_a' &= C_1+\left(\lambda_a-\mu_a\right)\left(1-\frac{w+a+m}{K}\right),\\ 
\mu_a' &= C_1,\\
\lambda_m' &= C_2+\left(\lambda_m-\mu_m\right)\left(1-\frac{w+a+m}{K}\right),\\ 
\mu_m' &= C_2,
\end{align*}
where $C_1, C_2>0$ are constants. % TODO what constants?

@@

We now assume that aneuploidy provides tolerance to drug therapy, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of wildtype cells, $\lambda_w - \mu_w < \lambda_a - \mu_a < 0$. We also assume that the decline are significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
We rewrite \cref{survival_prob} as
\begin{align*}
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
\text{and} \\
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) .
\end{align*}
Since $u,v\ll1$, this can be approximated by % TODO how do you do the first approx? taylor?
\begin{equation}\label{survprobwinitial}
\begin{aligned}
p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ %\label{survprobw2}
&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right]\\
%&=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right) .
\end{aligned}
\end{equation}
%Using the notational convention
%\begin{equation}\label{notationalconv}
%\Delta_i=\lambda_i-\mu_i,
%\end{equation}
%we write \eqref{survprobw} as
%\begin{equation}\label{survprobwapprox2}
%p_w=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right).
%\end{equation}

@@

We now assume that the growth rate of aneuploid cells is close to zero (either positive or negative), such that  $\left(\lambda_a-\mu_a-v\right)^2 < 4\lambda_avp_m$.
We rewrite \cref{survival_prob} as
\begin{equation}
p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a} .
\end{equation}
Using a following Taylor series expansion
\begin{equation*}
\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots,
\end{equation*}
we obtain the approximation
\begin{equation}
\begin{aligned}
p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
&=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
&=\frac{\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)^2+4\lambda_avp_m}{8\lambda_a\sqrt{\lambda_avp_m}}\\
&=\frac{4\lambda_avp_m+4\lambda_avp_m\left(1+\frac{\lambda_a-\mu_a-v}{2\sqrt{\lambda_avp_m}}\right)^2}{8\lambda_a\sqrt{\lambda_avp_m}}\\
&=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right).
\end{aligned}
\end{equation}
From \Cref{survprobwinitial}, % TODO I dont see how the referred equation plays a part here
the survival probability of a population starting from one wildtype individual is
\begin{equation}\label{scenario3}
\begin{aligned}
p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right].
%&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
\end{aligned}
\end{equation}

@@

To estimate the rescue probability $p_{rescue}$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
Thus, the rescue probability is given by 
\begin{equation}\label{rescue_prob}
p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w} ,
\end{equation}
where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.

Applying the approximations for the survival probability $p_w$ from \cref{survprobwapprox1,survprobwinitial,scenario3} and substituting $\Delta_k=\lambda_k-\mu_k$, we find that the rescue probability can be approximated by
\begin{equation}\label{rescue_prob_approx}
\begin{aligned}
&p_{rescue} \approx \\
  &\begin{cases}
  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
  \end{cases}
\end{aligned}
\end{equation}

@@

In \Cref{SurvPlot,SurvPlotLargeN} we explore the effects of the wildtype and aneuploid growth rates on the rescue probability for small and large population sizes ($N=10^4$ and $N=10^8$, respectively).

\Cref{SurvPlotNData} show $p_{rescue}$ as a function of $N$, including comparison of our first approximation \eqref{aneuploidyresistentfirstapprox} and simulation results.

\Cref{P_est,P_est_large_N} show the rescue probability for initial population sizes $N=10^4$ and $N=10^8$, respectively. % TODO check figure refs

In \Cref{DeleteriousPlot}, we compare the exact result (\cref{survival_prob}) with numerical simulations. The transition between the regimes defined by \cref{AneuploidyDeleteriousApprox} and \cref{pestthirdcase} respectively occurs at:
\begin{align}\label{thresholdvalueaneuploid}
\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}.
\end{align}

@@

In our analysis we used branching processes, which assume that growth (division and death) are density-independent. However, in some cases growth may be limited by resources (oxygen, nutrients, etc.) and therefore depend on cell density. 
We therefore performed stochastic simulations of a logistic growth model with carrying capacity $K$ (Methods). 
We find that ... (\Cref{rescue_prob_N_logistic}). % TODO check what we find; the figure just shows agreement of exact and stochastic

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Standing genetic variation}

In the above we assumed that upon beginning of drug therapy, the initial tumor consisted entirely of wildtype cells.
However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells REF), which are likely to have a deleterious effect (REF). But if the number of cells in the tumor $N$ is large, as expected if drug treatment is applied, there may already be a fraction $f=\tilde{u}/s$ of aneuploid cells in the population, where $s$ is the cost of aneuploidy (REF).

In this scenario, The probability of evolutionary rescue by cells with aneuploidy from the initial population is
\begin{equation*}
% redefine: p_total, p_denovo=p_resuce from eq above; p_stdvar is the new thing
p_{old} = 1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}. % TODO should be p_new?
\end{equation*}
The total probability of evolutionary rescue is given by
\begin{align}\nonumber
p_{total} 	&= p_{new}+\left(1-p_{new}\right)p_{old}\\ % TODO what is p_new?
			&= 1-\exp\left(-\left[\left(1-f\right)p_w + fp_a\right]N\right) .
\end{align}

The fraction of cases in which the population is rescued by the standing genetic variation is given by $F\left(f\right)=\frac{p_{old}}{p_{total}}$.
%=\frac{ 1-\e^{-fNp_a}}{1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}}.
Setting $F=\frac{1}{2}$, we use the expansion $\e^x \approx 1+x$ to obtain
\begin{equation}\label{halfeqstandvar} % TODO not sure this is interesting
f^*\approx\frac{p_w}{p_w+p_a}.
\end{equation}
% TODO maybe better to show p_total / p_denovo as function of u or N.
See \Cref{FractionPlot} for a demonstration of $F$ and $f*$.

@@

We find that the rescue ratio increase with the aneuploidy growth rate $\Delta_a$, because the better aneuploid cells are in growth, the better they are at rescuing the population (when they provide partial resistance) or delaying the extinction of the population (when they provide tolerance). 
However, the rescue decreases with the wildtype growth rate $\Delta_w$, because the better the wildtype is at growth, the less is depends on aneuploidy for rescue or delay, and the more likely it is to directly produce mutant cells, rather than relying on aneuploid cells for producing mutant cells (\Cref{rescue_ratio}).
The effect of the initial tumor size $N$ is mostly the same as the wildtype growth rate, except when aneuploidy provides tolerance and the population size is very large, in which case the ratio ... % TODO does H go to 1 in this case? is it lower than 1 for delta_a=-0.001? why?

@@

Assuming no aneuploidy ($u=0$), we define $T_1$ to be the time at which the first mutant cell appears that will avoid extinction and will therefore rescue the population. Note that if extinction occurs, that is $m_{t\rightarrow\infty}=0$, then it is implied that $T_1=\infty$, and vice versa if $T_1<\infty$ then $m_{t\rightarrow\infty}\neq0$.

The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = u p_a w_t$
where $w_t$ is the number of wildtype cells at time $t$, $w_t=N\e^{\Delta_w t}$.
The probability density function of $T_1$ is thus
$R\left(t\right)\e^{-\int_0^{t}{R(z)\d z}}$.
Therefore, the probability density function of $(T_1 \mid T_1 < \infty)$ is
$f_1(t) = \frac{R\left(t\right)\e^{-\int_0^{t} R(z)\d z}}{p_{rescue}}$. % Todo is this correct? I made changes

We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
\begin{align}\label{meantime1}
\tau_1 =
\int_{0}^{\infty}{t f_1(t) \d t} = 
\frac{\int_{0}^{\infty}{t u p_a N\e^{\Delta_w t} \e^{-\int_0^{t} u p_a N\e^{\Delta_w z} \d z} \d t}}{p_{rescue}} .
\end{align}
\begin{equation}
\frac{\int_0^\infty\e^{-\int_0^t R(z)\d z}\d t}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w t}-1}{\Delta_w}}\,\d t}{1-\left(1-p_w\right)^N}
\end{equation}
% YOAV HERE - dont understand how the previous two lines are the same...

>>>>

% line numbers
\usepackage[displaymath, mathlines]{lineno}
\renewcommand\linenumberfont{\normalfont\small\sffamily}
\linenumbers
\modulolinenumbers[2]

@@

@@

%%%%%%%%%%%
\begin{figure}[!t]
\centering
\includegraphics[width=0.75\textwidth]{Figures/figureAneuploidy.pdf}
\caption{
\textbf{Model illustration.}
\textbf{(A)} A population of cancer cells is composed of wildtype, aneuploid, and mutant cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively. 
Wildtype cells can become aneuploid at rate $u$. Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$. Color denotes the relative growth rates of the three genotypes such that $\lambda_w - \mu_w < \lambda_a - \mu_a < \lambda_m - \mu_m$. \textbf{(B)} The wildtype and the mutant are susceptible and resistant, respectively, to the drug. The aneuploid may be tolerant, non-growing, or or partially resistant.
}
\label{figureAneuploidy}
\end{figure}
%%%%%%%%%%%

@@

The state of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the state (\Cref{figureAneuploidy}):
\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\quad\left(\text{birth of wildtype cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
(-1,+1,0)&:\quad uw_t\quad\left(\text{wildtype cell becomes aneuploid}\right),\\
(-1,0,+1)&:\quad vw_t\quad\left(\text{wildtype cell becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,-1,+1)&:\quad va_t\quad\left(\text{aneuploid cell becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_am_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_am_t\quad\left(\text{death of mutant cell}\right).
\end{flalign*}
\end{subequations}
Each iteration of the simulation loop starts with computing the rates $\nu_j$ of each event $j$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $j$ is $p_j=\nu_j/\sum_i \nu_i$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
\begin{equation*}
m_t>\left\lfloor-\frac{3\log10}{\log\left(\frac{\mu_m}{\lambda_m}\right)}\right\rfloor+1,
\end{equation*}

\paragraph{$\tau$-leaping.}
When simulations are slow (e.g. due to large population size), we utilize $\tau$-leaping \citep{gillespie2001approximate}, where change in number of cells of genotype $i$ in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_i\Delta t$.
If the change in number of cells is negative and larger then the subpopulation size then the subpopulation size is updated to be zero.

@@

In our analysis we assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources.
A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
To simulate birth and death rates that depend on the number of cells in the population, we transform the rates of division and death to the following:
\begin{align*}
\lambda_w' &= \lambda_w, \\
\mu_w' &= \mu_w,\\
\lambda_a' &= C_1+\left(\lambda_a-\mu_a\right)\left(1-\frac{w+a+m}{K}\right),\\ 
\mu_a' &= C_1,\\
\lambda_m' &= C_2+\left(\lambda_m-\mu_m\right)\left(1-\frac{w+a+m}{K}\right),\\ 
\mu_m' &= C_2,
\end{align*}
where $C_1, C_2>0$ are constants. % TODO what constants? see Remus version.

@@

We now assume that aneuploidy provides tolerance to drug therapy, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of wildtype cells, $\lambda_w - \mu_w < \lambda_a - \mu_a < 0$. We also assume that the decline are significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
We rewrite \cref{survival_prob} as
\begin{equation}
\begin{aligned}
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
\text{and} \\
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) .
\end{aligned}
\end{equation}
Since $u,v\ll1$, the term in the root can be approximated using a 1st-order Taylor expansion. So, substituting the expressions for $p_a$ and $p_m$, we have
\begin{equation}\label{survprobwinitial}
\begin{aligned}
p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ %\label{survprobw2}
&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right]\\
%&=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right) .
\end{aligned}
\end{equation}

@@

We now assume that the growth rate of aneuploid cells is close to zero (either positive or negative), such that  $\left(\lambda_a-\mu_a-v\right)^2 < 4\lambda_avp_m$.
We rewrite \cref{survival_prob} as
\begin{equation}
p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a} .
\end{equation}
Using a following Taylor series expansion
\begin{equation*}
\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots,
\end{equation*}
we obtain the approximation
\begin{equation}
\begin{aligned}
p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
&=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
&=\frac{\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)^2+4\lambda_avp_m}{8\lambda_a\sqrt{\lambda_avp_m}}\\
&=\frac{4\lambda_avp_m+4\lambda_avp_m\left(1+\frac{\lambda_a-\mu_a-v}{2\sqrt{\lambda_avp_m}}\right)^2}{8\lambda_a\sqrt{\lambda_avp_m}}\\
&=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right).
\end{aligned}
\end{equation}
Plugging this in \cref{survprobwinitial}, the survival probability of a population starting from one wildtype individual is
\begin{equation}\label{scenario3}
\begin{aligned}
p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right].
%&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
\end{aligned}
\end{equation}

@@

To estimate the rescue probability $p_{rescue}$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
Thus, the rescue probability is given by 
\begin{subequations}
\begin{align}
p_{rescue} &= 1-\left(1-p_w\right)^N \label{rescue_prob} \\
& \approx 1-\e^{-Np_w} , \label{rescue_prob2}
\end{align}
\end{subequations}
where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.

Applying the approximations for the survival probability $p_w$ from \cref{survprobwapprox1,survprobwinitial,scenario3} in \cref{rescue_prob2} and substituting $\Delta_k=\lambda_k-\mu_k$, we find that the rescue probability can be approximated by
\begin{equation}\label{rescue_prob_approx}
\begin{aligned}
&p_{rescue} \approx \\
  &\begin{cases}
  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
  \end{cases}
\end{aligned}
\end{equation}

@@

@@

In our analysis we used branching processes, which assume that growth (division and death) are density-independent. However, growth may be limited by resources (oxygen, nutrients, etc.) and therefore depend on cell density. 
We therefore performed stochastic simulations of a logistic growth model with carrying capacity $K$ (Methods). 
We find that our approximations agree with results of simulations with density-dependent growth for biologically relevant parameter values (\Cref{rescue_prob_N_logistic}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Standing vs. de-novo genetic variation}

In the above we assumed that upon beginning of drug therapy, the initial tumor consisted entirely of wildtype cells.
However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells REF), which are likely to have a deleterious effect (REF). % TODO refs
But if the number of cells in the tumor $N$ is large, as expected if drug treatment is applied, there may already be a fraction $f=\tilde{u}/s$ of aneuploid cells in the population, where $s$ is the cost of aneuploidy (REF).

In this scenario, the probability of evolutionary rescue by cells with aneuploidy from the initial population is
\begin{equation*}
% redefine: p_total, p_denovo=p_resuce from eq above; p_stdvar is the new thing
p_{sgv} = 1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}. 
\end{equation*}
The total probability of evolutionary rescue is given by
\begin{align}\nonumber
p_{rescue} 	&= p_{sgv}+\left(1-p_{sgv}\right)p_{de-novo} \\
			&= 1-\exp\left(-\left[\left(1-f\right)p_w + fp_a\right]N\right) .
\end{align}

The fraction of cases in which the population is rescued by pre-existing aneuploid cells (i.e., standing genetic variation) is given by $F\left(f\right)=\frac{p_{sgv}}{p_{total}}$ (\Cref{FractionPlot}).

@@

We find that the rescue ratio increase with the aneuploidy growth rate $\Delta_a$, because the better aneuploid cells are in growth, the better they are at rescuing the population (when they provide partial resistance) or delaying the extinction of the population (when they provide tolerance). 
However, the rescue decreases with the wildtype growth rate $\Delta_w$, because the better the wildtype is at growth, the less is depends on aneuploidy for rescue or delay, and the more likely it is to directly produce mutant cells, rather than relying on aneuploid cells for producing mutant cells (\Cref{rescue_ratio}).
The effect of the initial tumor size $N$ is the similar to that of the wildtype growth rate. 
Importantly, in large tumors, the ratio converges to unity, that is, aneuploidy does not affect the probability for evolutionary rescue. % TODO can we find this analytical?

@@

Assuming no aneuploidy ($u=0$), we define $T_1$ to be the time at which the first mutant cell appears that will avoid extinction and will therefore rescue the population. Note that if extinction occurs, that is $m_{\infty}=0$, then it is implied that $T_1=\infty$, and vice versa if $T_1<\infty$ then $m_{\infty}>0$.

The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = u p_a w_t$,
where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
Note that 
\begin{equation}\label{integralR}
\int_0^{t}{R(z)\d z} = 
u p_a N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
u p_a N t,
\end{equation}
by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w}=t+O(t^2)$.
The probability density function of $T_1$ is thus
$R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)$. % TODO ref
Therefore, the probability density function of $(T_1 \mid T_1 < \infty)$ is
$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{p_{rescue}}$. 
\\

We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
\begin{equation}\label{meantime1}
\begin{aligned}
\tau_1 =
\int_{0}^{\infty}{t f_1(t) \d t} = 
\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}} = 
\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}}
\end{aligned}
\end{equation}
after applying integration by parts.
Therefore, plugging \cref{integralR,rescue_prob2} in \cref{meantime1}, 
\begin{align}\label{limitapprox}
\tau_1 =
\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx
\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u p_a N t} \d t =
\frac{1+\e^{-Np_w}}{u p_a N} ,
\end{align}
where we use the approximation $\left(1-\e^{-Np_w}\right)^{-1}\approx1+\e^{-Np_w}$ and performed the integration of the exponent.
% TODO continue from here
We plot the expansion \eqref{limitapprox} in Figure \ref{MeanTimeGrowthAneuploidyPlot} and observe that it is a very good fit for intermediate and large values of the initial wildtype population size.
